### 2010-07-29 Thesis/Chapter2.tex
- Update 29 July - Added Josh's corrections for first half of chapter 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@137 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Background in continuum mechanics for soft-tissue modelling}
\label{chap:softtissue}
\begin{shortAbstract}
As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation. In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations. If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial. Therefore, a substantial comprehension of the mechanics involved and a knowledge of the physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs. This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs. It will then present mathematical models able to describe the different mechanical aspects of materials and briefly discuss the mechanical characterisation of tissues.
\end{shortAbstract}

@@

Modelling anatomical structures requires an understanding of the deformation and stresses caused by the different sollicitations that occur during medical procedures. A sufficient knowledge of continuum mechanics is therefore essential to follow the rest of this manuscript. Continuum mechanics can be divided into two main parts: general principles common to all media (analysis of deformation, strain and stress concepts) and constitutive equations defining idealised materials. This chapter will not only deal with those two aspects but will also discuss the different aspects of the mechanical behaviour that we find in anatomical structures. Note that this chapter will mostly follow the notation used by \cite{Bonnet97} and \cite{Reddy07}. The interested reader may refer to these books for more details.  
%This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the mathematical models.

@@

Let us now consider a given particule of the body that we call $X$. What we will call particule in the following is in fact an infinitesimal volume of material. We denote the position it occupies in the initial configuration $\mathbf{X}$ and note $\mathbf{x}$ its position in the deformed configuration, both expressed in the chosen frame of reference. The mapping $\chi$ defined as the following:
\begin{equation}
\label{chap2:mapping}
\chi : \textbf{\textit{B}}_{\kappa_{0}} \rightarrow \textbf{\textit{B}}_{\kappa}
\end{equation} 
is called the \emph{deforming mapping} of the body $\textbf{\textit{B}}$ from $\kappa_{0}$ to $\kappa$. When analysing the deformation of a continuous body, it is necessary to describe the evolution of configurations through time. Its mathematical description follows one of the two approaches: the material description or the spatial description. The material description is known as \emph{Lagrangian description} whereas the spatial description is also called \emph{Eulerian description}. These two approaches are detailled next.

	\subsection{Lagrangian description}
In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$. Thus, the current coordinates ($\mathbf{x} \in \kappa$) are expressed in terms of the reference coordinates ($\mathbf{X} \in \kappa_{0}$): 
\begin{equation}
\textbf{x} = \chi(\textbf{X}, t), \quad \chi(\textbf{X}, 0) = \textbf{X},
\end{equation}
and the variation of a typical variable $\phi$ over the body is described with respect to the coordinates $\mathbf{X}$ and time $t$:
\begin{equation}
\phi = \phi(\textbf{X}, t).
\end{equation}
For a fixed value of $\mathbf{X} \in \kappa_{0}$, $\phi(\mathbf{X},t)$ gives the value of $\phi$ at time $t$ associated with the fixed particule $X$ whose position in the reference configuration is $\mathbf{X}$. We note that a change in time $t$ implies that the same particule $X$ has a different value $\phi$. Thus, the Lagrangian description focuses its attention on the particules of the continuous body and it is usually used in solid mechanics.

	\subsection{Eulerian description}
Rather than following the particules, the Eulerian description observes the changes at fixed locations. 	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\mathbf{x} \in \kappa$):
\begin{equation}
\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
\end{equation}
For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particule occupying the position $\textbf{x} \in \kappa$, which may very well be a different particule for each new time $ t $. Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particule, the Eulerian description is focused on a spatial position. This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}. Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.

	\subsection{Displacement field}
The displacement $ \mathbf{u} $ of a particule $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is given by the following:
\begin{equation}
\mathbf{u}(\textbf{X}, t) = \mathbf{x}(\mathbf{X}, t) - \mathbf{X}.
\end{equation}
A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body, which relates the deformed configuration $\kappa$ with the undeformed configuration $\kappa_{0}$. Indeed if the displacement field is known, we can construct the current configuration $\kappa$ from the undeformed configuration $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.

@@

Consider two particules $P_{1}$ and $P_{2}$ in a continuous body separated by an infinitesimal distance $d\mathbf{X}$:
\begin{equation}
\label{chap2:dX}
d\mathbf{X} = \mathbf{X}_{P_{2}} - \mathbf{X}_{P_{1}}.
\end{equation}
After deformation, the two particules have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
\begin{equation}
\mathbf{x}_{P_1} = \chi(\mathbf{X}_{P_{1}}, t) \quad \mbox{and} \quad \mathbf{x}_{P_2} = \chi(\mathbf{X}_{P_{2}}, t).
\end{equation}
Using \eqref{chap2:dX} the distance $d\textbf{x}$ between $P_{1}$ and $P_{2}$ can then be expressed as:
\begin{equation}
d\mathbf{x} = \mathbf{x}_{P_{2}} - \mathbf{x}_{P_{1}} = \chi(\mathbf{X}_{P_{1}} + d\mathbf{X}, t) - \chi(\mathbf{X}_{P_{1}}, t).
\end{equation}
The \emph{deformation gradient} $\textbf{F}$ can be defined as:
\begin{equation}
\mathbf{F} = \pd{\chi}{\mathbf{X}}
\end{equation}
and the vector $d\textbf{x}$ can then be obtained in terms of $d\mathbf{X}$ as:
\begin{equation}
\label{chap2:dxdX}
d\textbf{x} = \textbf{F} \, d\textbf{X}.
\end{equation}
Note that $\textbf{F}$ transforms vectors from the reference configuration into vectors in the current configuration and is therefore a second-order tensor.

@@

	\subsection{Change of surface}
For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration. Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation \TODO{Add figure with all notations}, the corresponding reference and current volumes are given respectively by:
\begin{align}
d\mathbf{V} &= d\mathbf{L} \cdot d\mathbf{A} \\
d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}
\end{align}
Using \eqref{chap2:dvdV} which relates the reference and deformed volumes and recalling that $d\mathbf{l} = \mathbf{F} \, d\mathbf{L}$, we have:
\begin{equation}
J \, d\mathbf{L} \cdot d\mathbf{A} = (\mathbf{F} \, d\mathbf{L}) \cdot d\mathbf{a}.
\end{equation}
The fact that this expression is valid for any vector $d\mathbf{L}$ and using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$ yields:
\begin{equation}
\label{chap2:dadA}
d\mathbf{a} = J \, \mathbf{F}^{-T} \, d\mathbf{A}.
\end{equation}

	\subsection{Volumetric and isochoric components}
We recall that the Jacobian $J$ has the physical meaning of being the local ratio of current to reference volume of a material volume element. Therefore, if $J = 1$ the volume does not change locally during the deformation and the latter is qualified as \emph{isochoric} at this given particule $P$. If $J = 1$ everywhere in the body, the deformation of the body is isochoric.

When dealing with incompressible and nearly incompressible materials it is necessary to separate the volumetric from the isochoric components of the deformation. Such a separation must ensure that the isochoric component $\mathbf{\hat{F}}$ does not imply any change in volume. This condition can be achieved by choosing $\mathbf{\hat{F}}$ as:
\begin{equation}
\mathbf{\hat{F}} = J^{-1/3} \mathbf{F}.
\end{equation}
Indeed,
\begin{align}
\det \mathbf{\hat{F}} &= \det (J^{-1/3} \mathbf{F}) \notag \\
&= (J^{-1/3})^3 \det \mathbf{F} \notag \\
&= J^{-1} J \notag \\
&= 1
\end{align}
The deformation gradient $\mathbf{F}$ can now be expressed in terms of the volumetric and isochoric component $J$ and $\mathbf{\hat{F}}$ as:
\begin{equation}
\mathbf{F} = J^{1/3} \mathbf{\hat{F}}
\end{equation}

@@

	\subsection{Cauchy-Green deformation tensors}
Let us consider two particules $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration. In the deformed configuration $P_{1}$ and $P_{2}$ occupy the positions $\tilde{P}_{1}$ and $\tilde{P}_{2}$ and they are separated by $d\mathbf{x}$. We are interested in the change of distance between the two points $P_{1}$ and $P_{2}$ as the body deforms.

@@

	\subsection{Green strain tensor}
Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
\begin{align}
(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
\end{align}
Let us define the \emph{Green-St.\,Venant strain tensor} $\mathbf{E}$ as:
\begin{equation}
\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
\end{equation}
so we can write:
\begin{equation}
(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
\end{equation}
By definition, the Green strain tensor is a symmetric second-order tensor. Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$. Using \eqref{chap2:gradient}, the Green strain tensor may be developped as the following:
\begin{align}
\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right] \label{chap2:GreenTensor}
\end{align}

The Green strain tensor can be expressed in terms of its components in any coordinate system. In particular, in the cartesian coordinate system $(X_1, X_2, X_3)$ the components $E_{ij}$ of $\mathbf{E}$ are the following:
\begin{equation}
E_{i,j} = \frac{1}{2} \left( \pd{u_i}{Xj} + \pd{u_j}{Xi} + \pd{u_k}{Xi} \pd{u_k}{Xj} \right), \quad i = 1, 2, 3.
\end{equation}
The components $E_{11}$, $E_{22}$ and $E_{33}$ are called \emph{normal strains} and it can be shown that they are in fact the ratio of the change in length to the original length along each of the three unit vectors. The components $E_{12}$, $E_{23}$ and $E_{13}$ are called \emph{shear strains} and they can be interpreted as a measure of the change in angle between line elements that were perpendicular to each other in the undeformed configuration.

@@

It can be shown that in the coordinate system $(\mathbf{p}_1, \mathbf{p}_2, \mathbf{p}_3)$ formed by the three eigenvectors, the expression of the strain tensor is:
\begin{equation}
\boldsymbol \varepsilon =
	\begin{bmatrix}
	\varepsilon_1 &          0           &           0           \\
	         0           & \varepsilon_2 &           0           \\
	         0           &           0           & \varepsilon_3
	\end{bmatrix}	
\end{equation}
Since there are no shear strain components in this particular coordinate system, the principal strains represent the maximum and minimum stretches of an elemental volume. Because of the obvious simplicity of the strain tensor's expression, this coordinate system is often used in mechanics to derive and solve equations.

@@

\section{Stress}
Stress is a measure of the intensity of the internal forces acting between particles of a deformable body across imaginary internal surfaces. These internal forces are produced between the particles in the body as a reaction to external forces applied on the body. The $SI$ unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square meter (unit area). As with strain, different measures of stress can be defined. For instance, stress can be measured per unit of deformed area (section \ref{chap2:CauchyStress}) or undeformed area (sections \ref{chap2:PiolaStress1} and \ref{chap2:PiolaStress2}). In general, the stress is not uniformly distributed over the cross section of a material body, and consequently the stress at a point on a given area is different from the average stress over this entire area. Therefore, it is necessary to define the stress not over a given area but at a specific point in the body. Stress at a point in a three-dimensional continuum can be measured in terms of nine quantities, three per plane, on three mutually perpendicular planes at the point. These nine quantities may be viewed as the components of a second-order tensor that we call \emph{stress tensor}.

@@

To establish this relationship, let us consider an infinitesimal tetrahedron in a Cartesian coordinate system. \TODO{Add figure of a tetrahedron with all notations}. We note $-\mathbf{t}_1$, $-\mathbf{t}_2$, $-\mathbf{t}_3$ and $\mathbf{t}$ the stress vectors in the outwards directions on the faces of the tetrahedron whose areas are respectively $\Delta a_1$, $\Delta a_2$, $\Delta a_3$ and $\Delta a$. For the mass $m$ inside the tetrahedron, we have by Newton's second law:
\begin{equation}
\label{chap2:Newton}
\sum \mathbf{F} = \mathbf{t} \Delta a - \mathbf{t}_1 \Delta a_1 - \mathbf{t}_2 \Delta a_2 - \mathbf{t}_3 \Delta a_3 + \Delta v \mathbf{f} = m \mathbf{a},
\end{equation}
where $\mathbf{f}$ is the force per unit volume acting on the body, $\Delta v$ the volume of the tetrahedron and $\mathbf{a}$ the acceleration.
The areas $\Delta a_1$, $\Delta a_2$ and $\Delta a_3$ being the projections of $\Delta a$, they are related to $\Delta a$ in the Cartesian coordinate system $(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ by:
\begin{equation}
\label{chap2:DeltaAi}
\Delta a_i = (\mathbf{n} \cdot \mathbf{e}_i) \Delta a, \quad i=1, 2, 3.
\end{equation}
Dividing \eqref{chap2:Newton} by $\Delta a$, using \eqref{chap2:DeltaAi} and noting that $\Delta v \slash \Delta a \rightarrow 0$ when the tetrahedron shrinks to a point yields:
\begin{align}
\mathbf{t} &= (\mathbf{n} \cdot \mathbf{e}_1) \mathbf{t}_1 + (\mathbf{n} \cdot \mathbf{e}_2) \mathbf{t}_2 + (\mathbf{n} \cdot \mathbf{e}_3) \mathbf{t}_3 \notag \\
&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3)
\end{align}

We define the \emph{Cauchy stress tensor} $\sigma = \mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3$ and we have:
\begin{equation}
\label{chap2:CauchyFormula}
\mathbf{t}(\mathbf{n}) = \mathbf{n} \cdot \sigma
\end{equation}
The stress vector $\mathbf{t}$ represents the vectorial stress on a plane whose normal is $\mathbf{n}$. As we demonstrated, the stress tensor is a property of the medium that is independent of $\mathbf{n}$. It represents the current force per unit of deformed area: $d\mathbf{f} = \mathbf{t} da = \sigma \cdot d\mathbf{a}$.

>>>>

\chapter{Background in continuum mechanics for soft-tissue modelling}
\label{chap:softtissue}
\begin{shortAbstract}
As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedures and per-operative guidance based on simulation. In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations. If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial. Therefore, a substantial comprehension of the mechanics involved and a knowledge of the physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs. This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs. It will then present mathematical models able to describe the different mechanical aspects of materials and briefly discuss the mechanical characterisation of tissues.
\end{shortAbstract}

@@

Modelling anatomical structures requires an understanding of the deformation and stresses caused by the different interactions that occur during medical procedures. A sufficient knowledge of continuum mechanics is therefore essential to follow the rest of this manuscript. Continuum mechanics can be divided into two main parts: general principles common to all media (analysis of deformation, strain and stress concepts) and constitutive equations defining idealised materials. This chapter will not only deal with those two aspects but will also discuss the different aspects of the mechanical behaviour that we find in anatomical structures. Note that this chapter will mostly follow the notation used by \cite{Bonnet97} and \cite{Reddy07}. The interested reader may refer to these books for more details.  
%This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the mathematical models.

@@

Let us now consider a given particle of the body that we call $X$. What we will call particle in the following is in fact an infinitesimal volume of material. We denote the position it occupies in the initial configuration $\mathbf{X}$ and note its position in the deformed configuration $\mathbf{x}$, both expressed in the chosen frame of reference. The mapping $\chi$ defined as the following:
\begin{equation}
\label{chap2:mapping}
\chi : \textbf{\textit{B}}_{\kappa_{0}} \rightarrow \textbf{\textit{B}}_{\kappa}
\end{equation} 
is called the \emph{deforming mapping} of the body $\textbf{\textit{B}}$ from $\kappa_{0}$ to $\kappa$. When analysing the deformation of a continuous body, it is necessary to describe the evolution of configurations through time. Its mathematical description follows one of the two approaches: the material description or the spatial description. The material description is known as \emph{Lagrangian description} whereas the spatial description is called \emph{Eulerian description}. These two approaches are detailed next.

	\subsection{Lagrangian description}
In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$. Thus, the current coordinates ($\mathbf{x} \in \kappa$) are expressed in terms of the reference coordinates ($\mathbf{X} \in \kappa_{0}$): 
\begin{equation}
\textbf{x} = \chi(\textbf{X}, t), \quad \chi(\textbf{X}, 0) = \textbf{X},
\end{equation}
and the variation of a typical variable $\phi$ over the body is described with respect to the coordinates $\mathbf{X}$ and time $t$:
\begin{equation}
\phi = \phi(\textbf{X}, t).
\end{equation}
For a fixed value of $\mathbf{X} \in \kappa_{0}$, $\phi(\mathbf{X},t)$ gives the value of $\phi$ at time $t$ associated with the fixed particle $X$ whose position in the reference configuration is $\mathbf{X}$. We note that a change in time $t$ implies that the same particle $X$ has a different value $\phi$. Thus, the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.

	\subsection{Eulerian description}
Rather than following the particles, the Eulerian description observes the changes at fixed locations. 	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\mathbf{x} \in \kappa$):
\begin{equation}
\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
\end{equation}
For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particle occupying the position $\textbf{x} \in \kappa$, which may very well be a different particle for each new time $ t $. Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particle, the Eulerian description is focused on a spatial position. This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}. Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.

	\subsection{Displacement field}
The displacement $ \mathbf{u} $ of a particle $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is given by the following:
\begin{equation}
\mathbf{u}(\textbf{X}, t) = \mathbf{x}(\mathbf{X}, t) - \mathbf{X}.
\end{equation}
A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body, which relates the deformed configuration $\kappa$ with the undeformed configuration $\kappa_{0}$. Indeed if the displacement field is known, we can construct the current configuration $\kappa$ from the undeformed configuration $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.

@@

Consider two particles $P_{1}$ and $P_{2}$ in a continuous body separated by an infinitesimal distance $d\mathbf{X}$:
\begin{equation}
\label{chap2:dX}
d\mathbf{X} = \mathbf{X}_{P_{2}} - \mathbf{X}_{P_{1}}.
\end{equation}
After deformation, the two particles have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
\begin{equation}
\mathbf{x}_{P_1} = \chi(\mathbf{X}_{P_{1}}, t) \quad \mbox{and} \quad \mathbf{x}_{P_2} = \chi(\mathbf{X}_{P_{2}}, t).
\end{equation}
Using \eqref{chap2:dX} the distance $d\textbf{x}$ between $P_{1}$ and $P_{2}$ can then be expressed as:
\begin{equation}
d\mathbf{x} = \mathbf{x}_{P_{2}} - \mathbf{x}_{P_{1}} = \chi(\mathbf{X}_{P_{1}} + d\mathbf{X}, t) - \chi(\mathbf{X}_{P_{1}}, t).
\end{equation}
The \emph{deformation gradient} $\textbf{F}$ can be defined as:
\begin{equation}
\mathbf{F} = \pd{\chi}{\mathbf{X}}
\end{equation}
and the vector $d\textbf{x}$ can then be obtained in terms of $d\mathbf{X}$ as:
\begin{equation}
\label{chap2:dxdX}
d\textbf{x} = \textbf{F} \, d\textbf{X}.
\end{equation}
Note that $\textbf{F}$ transforms vectors from the reference configuration into vectors in the current configuration and is therefore a second-order tensor.

@@

	\subsection{Change of surface}
For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration. Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation \TODO{Add figure with all notations}, the corresponding reference and current volumes are given respectively by:
\begin{align}
d\mathbf{V} &= d\mathbf{L} \cdot d\mathbf{A} \\
\mbox{and } \quad d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}.
\end{align}
Using \eqref{chap2:dvdV} which relates the reference and deformed volumes and recalling that $d\mathbf{l} = \mathbf{F} \, d\mathbf{L}$, we have:
\begin{equation}
J \, d\mathbf{L} \cdot d\mathbf{A} = (\mathbf{F} \, d\mathbf{L}) \cdot d\mathbf{a}.
\end{equation}
The fact that this expression is valid for any vector $d\mathbf{L}$ and using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$ yields:
\begin{equation}
\label{chap2:dadA}
d\mathbf{a} = J \, \mathbf{F}^{-T} \, d\mathbf{A}.
\end{equation}

	\subsection{Volumetric and isochoric components}
We recall that the Jacobian $J$ has the physical meaning of being the local ratio of current to reference volume of a material volume element. Therefore, if $J = 1$ the volume does not change locally during the deformation and the latter is qualified as \emph{isochoric} at this given particle $P$. If $J = 1$ everywhere in the body, the deformation of the body is isochoric.

When dealing with incompressible and nearly incompressible materials it is necessary to separate the volumetric from the isochoric components of the deformation. Such a separation must ensure that the isochoric component $\mathbf{\hat{F}}$ does not imply any change in volume. This condition can be achieved by choosing $\mathbf{\hat{F}}$ as:
\begin{equation}
\mathbf{\hat{F}} = J^{-1/3} \mathbf{F}.
\end{equation}
Indeed,
\begin{align}
\det \mathbf{\hat{F}} &= \det (J^{-1/3} \mathbf{F}) \notag \\
&= (J^{-1/3})^3 \det \mathbf{F} \notag \\
&= J^{-1} J \notag \\
&= 1.
\end{align}
The deformation gradient $\mathbf{F}$ can now be expressed in terms of the volumetric and isochoric component $J$ and $\mathbf{\hat{F}}$ as:
\begin{equation}
\mathbf{F} = J^{1/3} \mathbf{\hat{F}}.
\end{equation}

@@

	\subsection{Cauchy-Green deformation tensors}
Let us consider two particles $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration. In the deformed configuration $P_{1}$ and $P_{2}$ occupy the positions $\tilde{P}_{1}$ and $\tilde{P}_{2}$ and they are separated by $d\mathbf{x}$. We are interested in the change of distance between the two points $P_{1}$ and $P_{2}$ as the body deforms.

@@

	\subsection{Green strain tensor}
Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
\begin{align}
(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
\end{align}
Let us define the \emph{Green-St.\,Venant strain tensor} $\mathbf{E}$ as:
\begin{equation}
\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
\end{equation}
so we can write:
\begin{equation}
(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
\end{equation}
By definition, the Green strain tensor is a symmetric second-order tensor. Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$. Using \eqref{chap2:gradient}, the Green strain tensor may be expanded as the following:
\begin{align}
\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right]. \label{chap2:GreenTensor}
\end{align}

The Green strain tensor can be expressed in terms of its components in any coordinate system. In particular, in the Cartesian coordinate system $(X_1, X_2, X_3)$ the components $E_{ij}$ of $\mathbf{E}$ are the following:
\begin{equation}
E_{i,j} = \frac{1}{2} \left( \pd{u_i}{Xj} + \pd{u_j}{Xi} + \pd{u_k}{Xi} \pd{u_k}{Xj} \right), \quad i = 1, 2, 3.
\end{equation}
The components $E_{11}$, $E_{22}$ and $E_{33}$ are called \emph{normal strains} and it can be shown that they are in fact the ratio of the change in length to the original length along each of the three unit vectors. The components $E_{12}$, $E_{23}$ and $E_{13}$ are called \emph{shear strains} and they can be interpreted as a measure of the change in angle between line elements that were perpendicular to each other in the undeformed configuration.

@@

It can be shown that in the coordinate system $(\mathbf{p}_1, \mathbf{p}_2, \mathbf{p}_3)$ formed by the three eigenvectors, the expression of the strain tensor is:
\begin{equation}
\boldsymbol \varepsilon =
	\begin{bmatrix}
	\varepsilon_1 &          0           &           0           \\
	         0           & \varepsilon_2 &           0           \\
	         0           &           0           & \varepsilon_3
	\end{bmatrix}	
	.
\end{equation}
Since there are no shear strain components in this particular coordinate system, the principal strains represent the maximum and minimum stretches of an elemental volume. Because of the obvious simplicity of the strain tensor's expression, this coordinate system is often used in mechanics to derive and solve equations.

@@

\section{Stress}
Stress is a measure of the intensity of the internal forces acting between particles of a deformable body across imaginary internal surfaces. These internal forces are produced between the particles in the body as a reaction to external forces applied on the body. The $SI$ unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square metre (unit area). As with strain, different measures of stress can be defined. For instance, stress can be measured per unit of deformed area (section \ref{chap2:CauchyStress}) or undeformed area (sections \ref{chap2:PiolaStress1} and \ref{chap2:PiolaStress2}). In general, the stress is not uniformly distributed over the cross section of a material body, and consequently the stress at a point on a given area is different from the average stress over this entire area. Therefore, it is necessary to define the stress not over a given area but at a specific point in the body. Stress at a point in a three-dimensional continuum can be measured in terms of nine quantities, three per plane, on three mutually perpendicular planes at the point. These nine quantities may be viewed as the components of a second-order tensor that we call \emph{stress tensor}.

@@

To establish this relationship, let us consider an infinitesimal tetrahedron in a Cartesian coordinate system. \TODO{Add figure of a tetrahedron with all notations}. We note $-\mathbf{t}_1$, $-\mathbf{t}_2$, $-\mathbf{t}_3$ and $\mathbf{t}$ the stress vectors in the outwards directions on the faces of the tetrahedron whose areas are respectively $\Delta a_1$, $\Delta a_2$, $\Delta a_3$ and $\Delta a$. For the mass $m$ inside the tetrahedron, we have by Newton's second law:
\begin{equation}
\label{chap2:Newton}
\sum \mathbf{F} = \mathbf{t} \Delta a - \mathbf{t}_1 \Delta a_1 - \mathbf{t}_2 \Delta a_2 - \mathbf{t}_3 \Delta a_3 + \Delta v \mathbf{f} = m \mathbf{a},
\end{equation}
where $\mathbf{f}$ is the force per unit volume acting on the body, $\Delta v$ the volume of the tetrahedron and $\mathbf{a}$ the acceleration.
The areas $\Delta a_1$, $\Delta a_2$ and $\Delta a_3$ being the projections of $\Delta a$, they are related to $\Delta a$ in the Cartesian coordinate system $(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ by:
\begin{equation}
\label{chap2:DeltaAi}
\Delta a_i = (\mathbf{n} \cdot \mathbf{e}_i) \Delta a, \quad i=1, 2, 3.
\end{equation}
Dividing \eqref{chap2:Newton} by $\Delta a$, using \eqref{chap2:DeltaAi} and noting that $\Delta v \slash \Delta a \rightarrow 0$ when the tetrahedron shrinks to a point yields:
\begin{align}
\mathbf{t} &= (\mathbf{n} \cdot \mathbf{e}_1) \mathbf{t}_1 + (\mathbf{n} \cdot \mathbf{e}_2) \mathbf{t}_2 + (\mathbf{n} \cdot \mathbf{e}_3) \mathbf{t}_3 \notag \\
&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3).
\end{align}

We define the \emph{Cauchy stress tensor} $\sigma = \mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3$ and we have:
\begin{equation}
\label{chap2:CauchyFormula}
\mathbf{t}(\mathbf{n}) = \mathbf{n} \cdot \sigma .
\end{equation}
The stress vector $\mathbf{t}$ represents the vectorial stress on a plane whose normal is $\mathbf{n}$. As we demonstrated, the stress tensor is a property of the medium that is independent of $\mathbf{n}$. It represents the current force per unit of deformed area: $d\mathbf{f} = \mathbf{t} da = \sigma \cdot d\mathbf{a}$.
